---
title: "Spatial Packages"
author: "Ryan Peek"
date: "5/4/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial Mapping in R

Newer packages available (namely the `sf` package) make things much simpler and more streamlined for reading/writing and working with spatial data. A few nice writeups are [here](https://geographicdatascience.com/2017/01/06/first-impressions-from-sf-the-simple-features-r-package/) and [here](http://walkerke.github.io/2016/12/spatial-pipelines/), or [here](http://strimas.com/r/tidy-sf/). A handy CRS/spatial reference in R is [here](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)

First we'll need a few packages for this tutorial.

```{r packages, echo=T}
#devtools::install_github("tidyverse/ggplot2")
library(ggplot2) # for dev version
library(tidyverse) # piping and general tidying
library(sf) # rgdal/rgeos/sp replacement
library(viridis) #colors
library(leaflet)
library(rgdal)

# to view codes for lots of CRS, rgdal has these options:
#EPSG <- make_EPSG()
#EPSG[grep("aea", EPSG$note, ignore.case=TRUE), 1:2]

```

### Reading in Shapes

Turns out the `sf` package is much faster and more streamlined at reading in large spatial data. This is a set of **22,676** polygons with 20 fields total (about a 530MB shapefile). It takes just over 4 seconds with `sf`.

```{r readingshps_sf, eval=F, echo=T}
system.time(h12_sf <- st_read("~/Dropbox/R/PROJECTS/GIS_workshop_2017/data/Western_HUC12_update_cws.shp", stringsAsFactors = F))
as.tibble(h12_sf) 

```

Provides nice header with relevant info.

```
Reading layer `Western_HUC12_update_cws' from data source `/Users/ryanpeek/Dropbox/R/PROJECTS/GIS_workshop_2017/data/Western_HUC12_update_cws.shp' using driver `ESRI Shapefile'
converted into: POLYGON
Simple feature collection with 22676 features and 20 fields
geometry type:  MULTIPOLYGON
dimension:      XY
bbox:           xmin: -386100.3 ymin: -868185.9 xmax: 1132472 ymax: 1393179
epsg (SRID):    NA
proj4string:    +proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +datum=NAD83 +units=m +no_defs

   user  system elapsed 
  3.221   0.585   3.846
``` 
Object size is still big but smaller than `rgdal`:  

 - `object.size(h12_sf)`: *`504685024 bytes`*

You can also view the data using all the `tidy/dplyr` functions:

 - `glimpse(h12_sf) # see dataframe with list-col geometry col`

```
Observations: 22,676
Variables: 21
$ HUC_8      <chr> "16060004", "16050203", "16050202", "16040107"...
$ HUC_10     <chr> "1606000405", "1605020302", "1605020201", "160...
$ HUC_12     <chr> "160600040503", "160502030202", "160502020103"...
$ ACRES      <dbl> 19002, 25331, 24723, 39546, 14099, 14595, 2346...
$ NCONTRB_A  <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0...
$ HU_10_GNIS <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA...
$ HU_12_GNIS <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA...
$ HU_10_DS   <chr> "CLOSED BAS", "1605020303", "1605020202", "160...
$ HU_10_NAME <chr> "Rock Creek", "Bass Flat", "Town of Dayton-Car...
$ HU_10_MOD  <chr> "NM", "NM", "NM", "GC,IT", "NM", "NM", "NM", "...
$ HU_10_TYPE <chr> "C", "S", "S", "S", "C", "C", "S", "S", "S", "...
$ HU_12_DS   <chr> "160600040504", "160502030306", "160502020105"...
$ HU_12_NAME <chr> "Kingston Creek", "160502030202", "Gold Canyon...
$ HU_12_MOD  <chr> "NM", "NM", "UA,MA,TF,DM", "GC,IT", "NM", "NM"...
$ HU_12_TYPE <chr> "S", "S", "S", "S", "S", "S", "S", "S", "S", "...
$ META_ID    <chr> "NV01", "NV01", "NV01", "NV01", "NV01", "NV01"...
$ STATES     <chr> "NV", "NV", "NV", "NV", "NV", "NV", "NV", "NV"...
$ Shape_Leng <dbl> 44626.22, 67707.48, 64078.89, 63693.45, 38386....
$ Shape_Area <dbl> 76960949, 102545743, 100005384, 160166065, 570...
$ HUC_6      <chr> "160600", "160502", "160502", "160401", "16050...
$ geometry   <simple_feature> MULTIPOLYGON(((249516.13340..., MUL...
```

The same operation using traditional `sp::readOGR` takes a little over 16 seconds, and is a larger object in R.

```{r readingshps_sp, eval=F, echo=T}
library(rgdal)
dsn <- path.expand("~/Dropbox/R/PROJECTS/GIS_workshop_2017/data")
system.time(h12_sp <- readOGR(dsn = dsn, layer = "Western_HUC12_update_cws"))

object.size(h12_sp)
str(h12_sp) # lawd have mercy its S4
```
```
OGR data source with driver: ESRI Shapefile 
Source: "/Users/ryanpeek/Dropbox/R/PROJECTS/GIS_workshop_2017/data", layer: "Western_HUC12_update_cws"
with 22676 features
It has 20 fields

   user  system elapsed 
 13.256   2.741  16.126 
```
A slightly larger file size as well: 
 - `object.size(h12_sp)`: *559701824 bytes*

And importantly, the structure of the data is in the Large `SpatialPolygonsDataFrame` format, which is a bit trickier to work with in R.

```
Formal class 'SpatialPolygonsDataFrame' [package "sp"] with 5 slots
  ..@ data       :'data.frame':	22676 obs. of  20 variables:
  .. ..$ HUC_8     : Factor w/ 526 levels "15010001","15010002",..: 145 137 136 123 137 143 142 142 137 142 ...
  .. ..$ HUC_10    : Factor w/ 3913 levels "1501000101","1501000102",..: 1023 951 946 835 950 1001 993 992 952 992 ...
  .. ..$ HUC_12    : Factor w/ 20437 levels "150100010101",..: 6313 6020 6000 5495 6017 6226 6200 6195 6023 6199 ...
  .. ..$ ACRES     : num [1:22676] 19002 25331 24723 39546 14099 ...
  .. ..$ NCONTRB_A : num [1:22676] 0 0 0 0 0 0 0 0 0 0 ...
  .. ..$ HU_10_GNIS: Factor w/ 69 levels "10180 34025",..: NA NA NA NA NA NA NA NA NA NA ...
  .. ..$ HU_12_GNIS: Factor w/ 2003 levels "1001","10032 3953",..: NA NA NA NA NA NA NA NA NA NA ...
  .. ..$ HU_10_DS  : Factor w/ 1848 levels "0706010103","1206020204",..: 1846 490 488 438 1846 1846 509 508 491 508 ...
  .. ..$ HU_10_NAME: Factor w/ 3459 levels "1602030617-Frontal Bonneville Salt Flats",..: 2409 127 2893 568 865 1671 3027 759 396 759 ...
  .. ..$ HU_10_MOD : Factor w/ 674 levels "AD","AD,CD,PD",..: 567 567 567 212 567 567 567 567 567 567 ...
  .. ..$ HU_10_TYPE: Factor w/ 7 levels "C","F","I","M",..: 1 5 5 5 1 1 5 5 5 5 ...
  .. ..$ HU_12_DS  : Factor w/ 9720 levels "140102150105",..: 2954 2818 2808 2563 2814 2909 2896 2893 2818 2895 ...
  .. ..$ HU_12_NAME: Factor w/ 16667 levels "140500030714",..: 6683 257 5036 6732 10812 14678 5575 7615 258 7818 ...
  .. ..$ HU_12_MOD : Factor w/ 1091 levels "AD","AD,CD,SI,PD",..: 913 913 1088 388 913 913 388 913 913 913 ...
  .. ..$ HU_12_TYPE: Factor w/ 8 levels "C","F","I","M",..: 6 6 6 6 6 6 6 6 6 6 ...
  .. ..$ META_ID   : Factor w/ 60 levels "AZ01","AZ02",..: 32 32 32 32 32 32 32 32 32 32 ...
  .. ..$ STATES    : Factor w/ 47 levels "AZ","AZ,CA","AZ,CA,MX",..: 38 38 38 38 38 38 38 38 38 38 ...
  .. ..$ Shape_Leng: num [1:22676] 44626 67707 64079 63693 38387 ...
  .. ..$ Shape_Area: num [1:22676] 7.70e+07 1.03e+08 1.00e+08 1.60e+08 5.71e+07 ...
  .. ..$ HUC_6     : Factor w/ 72 levels "150100","150200",..: 27 25 25 22 25 27 27 27 25 27 ...
  ..@ polygons   :List of 22676
  .. ..$ :Formal class 'Polygons' [package "sp"] with 5 slots
  .. .. .. ..@ Polygons :List of 1
  .. .. .. .. ..$ :Formal class 'Polygon' [package "sp"] with 5 slots
  .. .. .. .. .. .. ..@ labpt  : num [1:2] 246070 140145
  .. .. .. .. .. .. ..@ area   : num 7.7e+07
  .. .. .. .. .. .. ..@ hole   : logi FALSE
  .. .. .. .. .. .. ..@ ringDir: int 1
  .. .. .. .. .. .. ..@ coords : num [1:1179, 1:2] 249516 249524 249538 249549 249552 ...
  .. .. .. ..@ plotOrder: int 1
  .. .. .. ..@ labpt    : num [1:2] 246070 140145
  .. .. .. ..@ ID       : chr "0"
  .. .. .. ..@ area     : num 7.7e+07
  .. ..$ :Formal class 'Polygons' [package "sp"] with 5 slots
  .. .. .. ..@ Polygons :List of 1
  .. .. .. .. ..$ :Formal class 'Polygon' [package "sp"] with 5 slots
  .. .. .. .. .. .. ..@ labpt  : num [1:2] 118003 137447
  .. .. .. .. .. .. ..@ area   : num 1.03e+08
  .. .. .. .. .. .. ..@ hole   : logi FALSE
  .. .. .. .. .. .. ..@ ringDir: int 1
  .. .. .. .. .. .. ..@ coords : num [1:4156, 1:2] 121273 121274 121276 121276 121276 ...
```

So for a half GB file that saves an immense amount of time, and it's a much cleaner way to work with the data.

### Looking at Bit Deeper at `sf` objects

It's possible to pull out just the list-column geometries to play with if that's something you need to do.

```{r diving into sf, echo=T}

h12_geom <- st_read("../data/shps/HUC8_named.shp")
(h12_geom <- st_geometry(h12_geom))
attributes(h12_geom)
h12_geom[[1]] %>% class

```

### Piping to `ggplot2`

Currently need the development version of ggplot2 for this, but it's awesome.

```{r ggplot}

#devtools::install_github("tidyverse/ggplot2", force=TRUE)
library(ggplot2)

rivs_sf <- st_read("../data/shps/major_rivers_dissolved.shp", stringsAsFactors = F, quiet = TRUE) %>% st_transform(3310)

st_crs(rivs_sf) # check crs
#"+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +datum=NAD83 +units=m +no_defs"

lakes_sf <- st_read("../data/shps/CA_major_lakes_res.shp", stringsAsFactors = T) %>% st_set_crs(3310)
st_crs(lakes_sf)

h8 <- st_read("../data/shps/HUC8_named.shp", stringsAsFactors = T) %>% st_transform(3310)
st_crs(h8)

# make a plot!
ggplot() +
  geom_sf(data = lakes_sf, aes(fill = FEATURE), show.legend = T) +   scale_fill_viridis("Feature Type", discrete = T) +
  geom_sf(data=rivs_sf, color="blue") +
  ggtitle("Lakes & Rivers CA") +
  theme_bw()

# This takes a long time (at least a few minutes)
ggplot() +
  geom_sf(data=h8, show.legend = F, color="gray") +
  ggtitle("HUC8 in CA") +
  theme_bw()

```


### Piping to Leaflet

You can also make nice leaflet maps using this package without too much hassle. 

```{r makeleaflet1}
library(leaflet)

rivers <- st_read("../data/shps/CA_major_rivers_CV_SNMdc.shp", stringsAsFactors = F) 
rivers <- st_transform(rivers, crs = 4326) # transform to WGS84
head(rivers)
st_proj_info("datum") # look at list of datums

rivers <- rivers %>% 
  as("Spatial")

rivers %>%
  leaflet() %>%
  addTiles() %>%
  addProviderTiles("Esri.WorldTopoMap") %>%
  addPolygons(weight = 1)

```

### Databases with `sf`

This is one of the more exciting bits, you can read/write/operate directly with spatial databases.

```{r read a db, echo=T}
fname <- file.path("~/Dropbox/gis/pisces.sqlite")
fname

# read layers in db:
st_layers(fname)
rivs <- st_read(fname, "major_rivers")
rivs <- st_transform(rivs, crs = 4326) # transform to WGS84
head(rivs)

rivs <- rivs %>% 
  as("Spatial")

# write to a shapefile:

# st_write(rivs, "../data/shps/major_rivers_pisces.shp")
```

### More Leaflet (fancy map)

```{r makeleaflet, echo=T}

# Make a Leaflet Map

leaflet() %>%
  addTiles() %>% 
  addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topo") %>%
  addPolylines(data=rivs, group="piscesRivs", weight=1) %>% 
  addPolylines(group = "BigRivs", stroke = 0.8,
               data = rivers, weight=1, color="maroon") %>% 
  
  addLayersControl(
    baseGroups = c("ESRI Aerial", "Topo"),
    overlayGroups = c("piscesRivs",
                      "BigRivs"),
    options = layersControlOptions(collapsed = T))


```

